### 2023-05-14 misc/expo/additive.tex
Second lemma of Roth and new book

>>>>

In the next lemma, we define the {\it diameter} of $X\subseteq \CC$ to be the quantity
$\sup_{z_1,z_2\in X} |z_1-z_2|$.

\proclaim Lemma {\advthm}. For every $\theta\in \RR$, $\eps>0$, and $n\in \NN$ sufficiently large, the
set $[n]$ can be partitioned
into disjoint arithmetic progressions $P_1, \ldots, P_m$, each of size at least $\sqrt{\eps n/(32 \pi)}$,
such that the diameter of $e(\theta P_i)\le \eps$ for all $1\le i\le m$.

\proof Let $t$ be a positive integer to be chosen later. Dividing the unit circle into $t$ segments of
length $2\pi/t$, by the pigeonhole principle there must be some segment that contains more than one of
the $t+1$ points $e(0), e(\theta), \ldots, e(t\theta)$. Thus there are $u$ and $v$ with $0\le u<v\le r$ such that
$\bigl|e(u\theta) - e(v\theta)\bigr|\le 2\pi/t$. Let $d = v-u$. By multiplying both $e(u\theta)$ and
$e(v\theta)$ by $e(-u\theta)$, we find that $\bigl|1-e(d\theta)\bigr| \le 2\pi/t$ as well. Then by telescoping and
applying the triangle inequality, we may expand
$$\bigl|1-e(sd\theta)\bigl| \le \sum_{k=1}^s \bigl|e\bigl((k-1)d\theta-e(kd\theta)\bigr|.$$
But by applying rotations as above, each of the terms in the sum can is equal to $\bigl|1-e(d\theta)\bigr|$,
implying that $\bigl|1-e(sd\theta)\bigl| \le 2\pi s/t$.

The reason this is important is that
whenever $P$ is any arithmetic progression with common difference $d$, any $x,y\in P$ will
have $\bigl|e(x\theta) - e(y\theta)\bigr| = \bigl|1-e(sd\theta)\bigr|$ for some $s$ that is at most
the length of $P$. Hence if $P$ is an arithmetic progression with common difference $d$ and length
at most $\eps t/(2\pi)$, then the diameter of $e(\theta P)$ will be at most $\eps$.

Take $[n]$ and partition it into residue classes modulo $d$. Within each residue class, we would like
to further group the elements into ``intervals'' of size roughly $\eps t/(2\pi)$. First we need to make
sure that the residue classes are large enough. If a residue class has size at least $\eps t/(2\pi)$,
then it can be subdivided into intervals of size between $\eps t/(4\pi)$ and $\eps t/(2\pi)$.
To ensure this, we note that the residue classes have size $\lfloor n/d\rfloor \ge \lfloor n/t\rfloor$,
so we shall set $t \le \sqrt{2\pi n / \eps}$. With this choice we find that each $P_i$ has size
$$\biggl\lfloor {\eps t\over 4\pi}\biggr\rfloor \le  \Biggl\lfloor {\eps\over 4\pi} \sqrt{2\pi n\over \eps}
\Biggr\rfloor = \Biggl\lfloor \sqrt{\eps n\over 8\pi} \Biggr\rfloor \le \sqrt{\eps n\over 32\pi}.$$
